Lossy compression of DNA sequencing quality data

### 2015-02-06 manuscript/q-compression.tex
More polishing of the intro.

Read data from high-throughput sequencing constitutes the largest
category of data in genomics research because of great redundancy,
inclusion of quality values, and read-level naming and
metadata. Because of this abundance effective compression of read data
has the potential for substantial improvement in data storage and
transfer efficiency. Among the several challenges for compression of
read data is dealing with different error profiles resulting from
differences in underlying chemistries, signal detection and processing
mechanisms, inherent biases, and other idiosyncratic properties of
distinct high-throughput sequencing technologies. Sequence reads
generated by instruments such as an Illumina HiSeq, the focus of this
research, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}).

Quality values comprise a standard component of \textsc{fastq} format
files~\cite{Cock:2010ve}.  At the level of sequence read the
probability of error for each base-call is typically represented by
\textsc{phred} quality value, which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). These errors are represented a single quality value per
base-call for Illumina sequence reads.

Quality values are typically used throughout bioinformatics pipelines.
Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, the elimination of reads that on the whole do
not meet quality standards, which reduces the total number of reads;
and \textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between query and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
with high-quality calls are deemed more reliable than those with
low-quality calls, particularly in low coverage
regions~\cite[e.g.,][]{McKenna:2010bh}.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
Although some recent work has included, or focused on, compression of
quality
values~\cite[e.g.,][]{Tembe:2010ys,Kozanitis:2011kl,asnani2012lossy,Hach:2012ys,Wan:2012kq,janin2013adaptive,Ochoa:2013rt,Canovas:2014fr,zhou2014compression,DBLP:conf/recomb/YuYB14}.
Typical compression methods fail to utilize the biases inherent in
existing sequencing technologies. Depending on the sequencing
technology used, the base calling is more accurate in the earlier
cycles, gradually losing quality in later
cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some absolute loss of
information is deemed acceptable if the inferences from analyses are
relatively unaffected. Therefore we here describe our research
investigating some general strategies for compression of sequence read
quality values. Our objective provide perspective on several
strategies logically applied to Illumina sequence read quality values
rather than to generate a robust high-quality software for
use. Consistent with our perspective we place emphasis not only on the
compression efficiency, but also the effect of lossy compression on
typical subsequent analyses.  We examine three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
consider an exemplar of each class.

Lossy compression compromises representation accuracy for better space
efficiency. Different lossy compression methods have different
properties, therefore we need to evaluate a number of tradeoffs for
each of our proposed methods. More importantly, we need to examine
each approach in the context of typical bioinformatics tasks, such as
quality control, \emph{de novo} genome assembly, and aligning the
sequences to a given reference.

One compression strategy is to look at the distribution of quality
values and see if we can summarize the information contained within
ranges of values. Specifically, we can reduce the dimensionality of
the distribution of quality values by binning together adjacent
quality values. We can select the number of bins depending on the
desired amount of compression and retention of quality
information. Taken to nearly the extreme, a base call can be
considered either good or bad. This is the motivation of our approach,
deemed quality value binning, where we observe the distribution of
quality values across the read set and find a cutoff between good and
bad bases.

Another compression strategy involves simplifying the data through
modeling of the quality values of a given sequence. By constructing a
compact model of our quality values, we only have to store the
parameters of the model to recreate the quality sequences.
Intuitively, the simpler the model, the better the compression, but
results in potentially higher differences compared to the original
quality sequences. Another benefit of the model-based compression is
the ability to decompress any particular quality value in the sequence
with constant time during decompression. In our approach, we represent
the quality values of a sequence as a polynomial function, but it is
possible to use other models (such as other basis functions or
splines).

Profile-based compression utilizes the fact that in a vast set of
quality sequences one can identify large classes of highly similar
elements. If one could partition the data into such classes, each of
those would be efficiently representable with a single consensus
profile.

@@

The economic costs of memory use for binning, in general terms,
include no fixed costs, and marginal costs that are a function of the
number of base-call quality values encoded times the cost of the
encoding.

@@

\section{Some Alternative Text}

The following text is taken from a draft grant proposal, and the
content overlaps considerably with the other text in this
document. However, at least in some cases the wording of this
alternative text might be preferable and serve as a replacement. In
any event the entire draft manuscript has to be substantially revised.

\subsubsection{Pilot Study: profile-based compression of quality 
values}\label{sec:profile-based}

Quality value patterns can be readily determined by clustering
sequences of quality values and using the resulting cluster centers as
representative profiles. In our pilot study we used $k$-means
clustering, a vector quantization method that partitions a set of
samples into $k$ sets that minimize within-cluster sum of
squares~\cite{hastie_09_elements-of.statistical-learning}, as follows.

\begin{enumerate}
\item
Randomly sample an arbitrary number of quality sequences as a training
set.
\item
Perform $k$-means clustering on the sample, and use the cluster
centers as profiles for the dataset.
\item
Replace each quality value sequence with the index value of the
profile closest in Euclidean distance.
\end{enumerate}

Our pilot study results demonstrate that $k$ values as small as
32--256 sufficiently represent patterns in Illumina read quality
values such that performance in subsequent analyses is very close to
uncompressed data (see Section \ref{sec:pilot-evaluation}). The
resulting compressed quality component of the original set of
\textsc{fastq} files consists of the $k$ quality profiles (the fixed
size cost), and a single byte representing the assigned quality
profile index for each read (the marginal size cost). Therefore the
compression factor is proportional of read length: the marginal size
cost of 1 byte/read for Illumina HiSeq read lengths of 100 bp requires
0.01 bits/base-call, and MiSeq read lengths of 300 bp require 0.0033
bits/base-call. The fixed cost of profiles of course reduces the
compression ratio, but this fixed cost is amortized across reads in a
set. Both fixed and marginal costs can be further reduced with
subsequent compression of the index and profiles (e.g., using general
compression algorithms). Our pilot study demonstrates that using 256
profiles with subsequent bzip2 compression (using a Burrows-Wheeler
block-sorting compression algorithm, and Huffman coding) represents
quality values of Illumina 100 bp reads at 0.074 bits/base-call
compared to the original data compressed with bzip2 only of 2.11
bits/base-call for the fragment library and 2.70 bits/base-call for
the short jump library. The very recently published
Read-Quality-Sparsifier
(\textsc{rqs};~\cite{DBLP:conf/recomb/YuYB14}), which apparently
represents the current state-of-the-art for compressing quality
values, achieved best case compression of 0.254 bits/base-call, and a
mean compression of 1.841 bits/base-call. Whereas our profile method
achieves $\sim$3.432$\times$ more compression than textsc{rqs} best
performance, and $\sim$24.878$\times$ more than average
performance. Note too that the compression ratio for \textsc{rqs}
varies with the $k$-mer composition of the data, whereas profile-based
compression is invariant to $k$-mer composition.

\subsubsection{Evaluating compression effects on subsequent 
analysis results}\label{sec:pilot-evaluation}

Lossy compression compromises representation accuracy for better space
efficiency. Different lossy compression methods have different
properties, therefore we need to evaluate performance in a number of
analysis contexts, and use multiple performance measures to
appropriately evaluate lossy compression methods. Important
bioinformatics contexts include typical tasks, such as quality control
processing, \textit{de novo} genome assembly, and aligning reads to a
reference. In our pilot study we again used Illumina reads from
fragment and short-jump libraries of the \textit{Rhodobacter
  sphaeroides} as test data, and assessed the effects of lossy
compression on each of the aforementioned bioinformatics
operations. The results for the profile-based compression follow.

\paragraph{Aligning reads to a reference sequence}
Another important analysis involving read data are the alignments of
reads to a reference sequence. Such alignments are performed for
genome annotation, quantification of gene expression in RNA-Seq
studies, estimation of sequencing coverage, and other purposes. To
evaluate the effect of compression on read alignment we used
\textsc{bowtie} 2~\cite{Langmead:2012rw} with default settings and
max- and min-mismatch parameters of \texttt{-mp 6,6} and \texttt{-mp
  2,2} respectively. The proportion of reads mapped using
compressed/decompressed data was extremely close to that of the
uncompressed data: 0.999 for the fragment library reads, and 0.998 for
the short-jump library reads.

>>>>

Read data from high-throughput sequencing constitutes the largest
category of data in genomics research because of great redundancy,
inclusion of quality values, and read-level naming and
metadata. Because of this abundance effective compression of read data
has the potential for substantial improvement in data storage and
transfer efficiency.

Quality values comprise a standard component of \textsc{fastq}
files~\cite{Cock:2010ve}, a very common format for sequence read data.
At the level of sequence read the probability of error for each
base-call is typically represented by \textsc{phred} quality value,
which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). There is a single quality value per base-call for Illumina
sequence reads.

Quality values can be used throughout bioinformatics pipelines.  Among
the most fundamental uses of sequence quality values is as part of the
quality assessment and quality control (\textsc{qa/qc}) processes
prior to subsequent analysis steps. Quality control based on quality
values generally includes two operations: \textit{i}.~filtering, the
elimination of reads that on the whole do not meet arbitrary quality
standards, which reduces the total number of reads; and
\textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in low
coverage regions.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr},
although some recent work has addressed compression of quality
values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}.
Among the several challenges for compression of read data is dealing
with different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Sequence reads generated by instruments such
as an Illumina HiSeq, the focus of this research, are characterized by
having relatively few insertion and deletion errors, but substitution
(miscall) errors are much more frequent and have context-specific
patterns. These errors are non-uniformly distributed over the read
length (e.g., error rates increase up to $\sim$16$\times$ at the
3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low quality tails
at the 3$^{\prime}$ end~\cite{Minoche:2011km}).

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some loss of information is
deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop a robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and consider an exemplar of each class. In addition to
quantifying compression efficiency, we also assess the effects of
quality value information loss resulting from compression on
subsequent genomic analyses including read preprocessing (filtering
and trimming), genome assembly, and read mapping.

@@

The economic costs of memory use for binning, in general terms,
include no fixed costs, and marginal costs that are a function of the
number of base-call quality values times the cost of the encoding.

@@
